Instant Continuation Marks--Just Add $\lambda$

### 2012-10-01 paper.tex
In the middle of more points.

Numerous programming language instruments rely on stack inspection to function. Statistical profilers sample the stack regularly to record active functions, algebraic steppers observe the stack to represent the evaluation context of an expression, and debuggers naturally require consistent access to the stack. Each scenario relies on implementation-specific information and must be maintained as the instrumented language undergoes optimizations and ventures across platforms. This makes these tools brittle and increases the porting cost of the language ecosystem. Each scenario would benefit from a generalized stack-inspection mechanism available within the instrumented language itself. If written in such an enhanced language, each instrument would be more robust, more easily modified, and would port for free.

@@

We begin by explaining continuation marks further. We then present a core and enhanced language along with Redex models for each. We discuss the necessary properties for a meaningful transform.  We present two transformations of continuation marks--one direct and one CPS-like--to the call-by-value \lc. We then prove a meaning-preservation theorem.

@@

The evaluation of the factorial of 2, here expressed by \scheme|(fac 2 1)| proceeds as
\begin{schemedisplay}
(fac 2 1)
(if (zero? 2) ... (wcm 2 (fac (- 2 1) (* 2 1))))
(wcm 2 (fac (- 2 1) (* 2 1)))
(wcm 2 (fac 1 (* 2 1)))
(wcm 2 (fac 1 2))
(wcm 2 (if (zero? 1) ... (wcm 1 (fac (- 1 1) (* 1 2)))))
(wcm 2 (wcm 1 (fac (- 1 1) (* 1 2))))
\end{schemedisplay}
At this point, there is one \scheme'wcm' directive directly nested within another. Behaviorally, the outer one is collapsed \emph{before} evaluation of the inner body proceeds. In effect, it is overwritten. Following this, evaluation continues as
\begin{schemedisplay}
(wcm 1 (fac (- 1 1) (* 1 2)))
(wcm 1 (fac 0 (* 1 2)))
(wcm 1 (fac 0 2))
(wcm 1 (if (zero? 0) (begin (print (ccm)) 2) ...))
(wcm 1 (begin (print (ccm)) 2))
(wcm 1 (begin (print '(1)) 2))
(wcm 1 (begin 2))
(wcm 1 2)
2
\end{schemedisplay}
The visible effect of the call is
\begin{verbatim}
> (fac 2 1)
(1)
2
\end{verbatim}
and the influence of the tail-call behavior is evident.

%Language transformations must be meaningful: they must preserve meaning or provide it. Continuation marks have special meaning in the presence of tail-call optimization.
% uses of language transformation:
% meaning of one language in terms of another (desugaring)
% - reduce core code which simplifies implementations
% easier to reason/optimize

%Language transformations are used for many reasons, but they he transformation must be meaningful: it must either preserve meaning or provide it.

%Language transformations either provide meaning or preserve meaning.

%Language transformation is useful for a variety of purposes. It can be used to define one language in terms of another, It can be used to give meaning to the source language in terms of the target language [cite something], provide a characterization more amenable to analysis or optimization. Language transformation can also be used to \emph{desugar} a language into a simpler core. This 
%Transforming has the following benefits:
%reduce core code (desugaring, simplifies implementations)
%give meaning to constructs (semantically, as well as, say, macros)
%local transformations - macros, for example
%global transformations - compilers of all kinds, for example

%talk about the common practice of analyzing an extension of the lambda calculus, and a subpractice of doing so via a transformation to the core.

%While continuation marks have had significant treatment, they do not currently enjoy a transformation to the plain $\lambda$-calculus.

%\section{Continuation Marks}

%Continuation marks generalize stack inspection, allowing a bevy of features that rely on some form of it to be defined at the language level. This includes algebraic steppers \cite{clements2001modeling}, security policies \cite{clements2004tail}, and aspects \cite{tucker2003pointcuts}. By defining such tools at the language level, such tools are more robust, cheaper to port, more easily changed, etc.

%Despite their profound utility, few languages offer continuation marks. Their inception lays in Racket \cite{plt-tr1} and they have been added experimentally to \emph{JavaScript} \cite{clements2008implementing}, but their earnest adoption by other languages has been nonexistent.

% transformation means that they can be compiled away and any higher order language can have them
% transformation also means that they do not need to be built into the language

% talk about composability of this transform to others

%Racket offers desugaring facilities to the user via a powerful macro system. This requires and is limited to local transformations limited to expression boundaries.

%A transformation-based compiler, such as a CPS compiler \cite{appel2007compiling}, effects a global transformation of the program. This effect discards original source or transforms it so radically that it is suitable only for the levels beneath the programmer's regular interaction.

@@

We begin by formally defining the core language the transformation produces and the extended language over which the transformation is defined.

@@

The target language of the transformation, \lv, is simply the call-by-value \lc.

@@

The source language of the transformation, \cm, is an extension of \lv\ with facilities for continuation marks.

@@

The sole novelty of \cm\ above the plain \lv\ is the ability to annotate and observe the continuation. Our evaluation model of \cm, given by Pettyjohn et al. \cite{pettyjohn2005continuations}, utilizes the current evaluation context for that purpose. The mutually inductive valuation contexts $E$, $F$ are defined by
\begin{align}
E=\, &\wcm{v}{F}\,|\,F\\
F=\, &\app{E}{e}\,|\,\app{v}{E}\,|\,\wcm{E}{e}\,|\,\hole
\end{align}
where $\hole$ denotes a ``hole'' in the evaluation context, the ultimate destination of the evaluation of the expression that previously resided there.

The semantics of \cm\ are
\begin{align}
E[\app{\abs{x}{e}}{v}]  &\rr E[e[x\leftarrow v]]\label{reduction}\\
E[\wcm{v}{\wcm{v'}{e}}] &\rr E[\wcm{v'}{e}]\label{tail-enforce}\\
E[\wcm{v}{v'}]          &\rr E[v']\label{wcm-body}\\
E[\ccm]                 &\rr E[\chi(E)]\label{ccm-in-context}
\end{align}
where $\chi$ is defined by
\begin{equation}
\chi(E)=\chi'(E,\mathbf{nil})
\end{equation}
and $\chi'$ is defined in turn by
\begin{align}
\chi'(\wcm{v}{F},vs) &= \chi'(F,\abs{z}{\app{\app{z}{v}}{vs}})\\
\chi'(\app{E}{e},vs) &= \chi'(E,vs)\\
\chi'(\app{v}{E},vs) &= \chi'(E,vs)\\
\chi'(\wcm{E}{e},vs) &= \chi'(E,vs)\\
\chi'(\hole,vs)      &= vs
\end{align}

As an extension of \lv, \cm\ inherits rule \ref{reduction} directly. Rule \ref{tail-enforce} expresses the tail-call behavior of \cm\ in that a new mark made in tail position overwrites the previous mark. Rule \ref{wcm-body} defines the value of a \scheme'wcm' expression to take on the value of the body. Finally, rule \ref{ccm-in-context} defines the value of a \scheme'ccm' expression in terms of its context via the metafunction $\chi$. The value of the function is a list of the current continuation marks.

Notice that $\chi'$ is defined in accumulator-passing style which has the effect of ordering the marks from most- to least-recently placed. The direct style formulation reverses this order. In the end, the ordering chosen is immaterial; we use this formulation to make the transform simpler.

@@

Language transformations must be meaningful: they must preserve meaning or provide it. As transformations, desugarings put the meaning of one construct in terms of another. If the meaning of the former was defined only informally, the desugaring offers a new definition as formal as the definition of the latter. In this way, desugarings offer a convenient way to provide meaning to a construct and avoid enlarging the language. If, however, the sugared language already has a formal meaning, the desugaring must preserve it. This will be made more precise shortly.

The transformation $\mathcal{C}$ is defined over each syntactic form of \cm. In order to preserve the meaning of \cm, $\mathcal{C}$ must commute with evaluation. More precisely, the following property should hold for programs $p\in\lambda_{cm}$:
\[
\begin{array}{ccc}
p & \cmrrs & v\\
\downarrow_\mathcal{C} & & \downarrow_\mathcal{C}\\
\C{p} & \lvrrs & \C{v}
\end{array}
\]
or, succinctly,
\begin{equation}
\label{meaning-preservation-property}
p\cmrrs v\implies\C{p}\lvrrs\C{v}
\end{equation}

%Our first approach was to use CPS because it treats all data and control flow uniformly [cite Sabry]. We passed an additional parameter (or two) in the spirit of double-barrelled continuations (but only in spirit). Our first attempt passed a pair as the second argument that needed to be deconstructed and constructed in every recursive definition. This was a lot of mindless manipulation that had little to do with the essence of the transform. In what took far too long a time, we realized the extent of this manipulation and finally confronted the subconscious aversion to passing \emph{two} arguments which we found was baseless.

%Almost immediately after, we realized that CPS was wholly unnecessary and landed back where we started with a direct-style definition.

%A comprehensive language transformation need only be defined over each syntactic form of the language. Thus, our full transformation definition will comprise transformation definitions for each of the five syntactic forms of \cm.

As variables, abstractions, and applications exist in the target language, their transformation is relatively straightforward. The essence of \cm\ is that programs can apply information to and observe information about the context in which they are evaluated. This characterization makes the work of the transform clear: to propagate contextual information to each expression, manipulating it appropriately as the context changes.

>>>>

Numerous programming language instruments rely on stack inspection to function. Statistical profilers sample the stack regularly to record active functions, algebraic steppers observe the stack to represent the evaluation context of an expression, and debuggers naturally require consistent access to the stack. In each of these scenarios, the instrument relies on implementation-specific information and must be maintained as the instrumented language undergoes optimizations and ventures across platforms. This makes these instruments brittle and increases the porting cost of the language ecosystem. Each would benefit from a generalized stack-inspection mechanism available within the instrumented language itself. If written in such an enhanced language, each instrument would be more robust, more easily modified, and would port for free.

@@

We begin by further explaining continuation marks. We then formalize them via the introduction of a core and enhanced language with Redex models for each. We discuss the necessary properties for a meaningful transform.  We present two transformations of continuation marks--one direct and one CPS-like--to the call-by-value \lc. We then prove a meaning-preservation theorem.

@@

The evaluation of the factorial of 2, here expressed by \scheme|(fac 2 1)|, proceeds as
\begin{schemedisplay}
(fac 2 1)
(if (zero? 2) ... (wcm 2 (fac (- 2 1) (* 2 1))))
(wcm 2 (fac (- 2 1) (* 2 1)))
(wcm 2 (fac 1 (* 2 1)))
(wcm 2 (fac 1 2))
(wcm 2 (if (zero? 1) ... (wcm 1 (fac (- 1 1) (* 1 2)))))
(wcm 2 (wcm 1 (fac (- 1 1) (* 1 2))))
\end{schemedisplay}
At this point, there is one \scheme'wcm' directive directly nested within another. Behaviorally, the outer one is collapsed \emph{before} evaluation of the inner body proceeds. In effect, it is overwritten. Following this, evaluation continues as
\begin{schemedisplay}
(wcm 1 (fac (- 1 1) (* 1 2)))
(wcm 1 (fac 0 (* 1 2)))
(wcm 1 (fac 0 2))
(wcm 1 (if (zero? 0) (begin (print (ccm)) 2) ...))
(wcm 1 (begin (print (ccm)) 2))
(wcm 1 (begin (print '(1)) 2))
(wcm 1 (begin 2))
(wcm 1 2)
2
\end{schemedisplay}
The visible effect of the call is
\begin{verbatim}
> (fac 2 1)
(1)
2
\end{verbatim}
from which the overwriting which occurred is evidenced the singleton list obtained by \scheme|(ccm)|.

Using this mechanism, the principal onus of a statistical profiler, algebraic stepper, or debugger can be met by a straightforward language transformation which recursively wraps each term in a \scheme'wcm' directive annotating the continuation with a description of the wrapped term.

With this understanding of continuation marks, we can now formalize them behaviorally.

@@

In order to formalize continuation marks, we introduce an extension of the \lc with facilities to manipulate continuation marks and present its semantics in the style of Felleisen and Hieb \cite{felleisen1992revised}. Because our ultimate goal is a desugaring transform, we first introduce the target language of the transform, the \lc, in the same way.

@@

The target language of the transform is Plotkin's call-by-value \lc, \lv\ \cite{plotkin1975call}.

@@

The source language of the transformation is an extension of \lv\ with facilities for continuation marks which we term \cm.

@@

The sole novelty of \cm\ above the plain \lv\ is the ability to annotate and observe the continuation. Our evaluation model of \cm, given by Pettyjohn et al. \cite{pettyjohn2005continuations}, utilizes the current evaluation context for that purpose. Evaluation contexts $E$ are defined by
\begin{align}
E=\, &\wcm{v}{F}\,|\,F\\
F=\, &\app{E}{e}\,|\,\app{v}{E}\,|\,\wcm{E}{e}\,|\,\hole
\end{align}
The definition of $E$ is crafted to prevent directly nested \scheme'wcm' directives from occurring in valid evaluation contexts.

The semantics of \cm\ are
\begin{align}
E[\app{\abs{x}{e}}{v}]  &\rr E[e[x\leftarrow v]]\label{reduction}\\
E[\wcm{v}{\wcm{v'}{e}}] &\rr E[\wcm{v'}{e}]\label{tail-enforce}\\
E[\wcm{v}{v'}]          &\rr E[v']\label{wcm-body}\\
E[\ccm]                 &\rr E[\chi(E)]\label{ccm-in-context}
\end{align}

As an extension of \lv, \cm\ inherits rule \ref{reduction} with substitution carried through \scheme'wcm' terms.

If directly-nested \scheme'wcm' directives are introduced into the evaluation context, invalidating it (as in the tail-recursive factorial example), rule \ref{tail-enforce} collapses the outer into the inner.

Rule \ref{wcm-body} defines the value of a \scheme'wcm' expression to take on the value of the body, once obtained.

The definition of rule \ref{ccm-in-context} is given in terms of a metafunction $\chi$ defined by
\begin{equation}
\chi(E)=\chi'(E,\abs{x}{\abs{y}{y}})
\end{equation}
where $\chi'$ is defined by
\begin{align}
\chi'(\wcm{v}{F},vs) &= \chi'(F,\abs{z}{\app{\app{z}{v}}{vs}})\\
\chi'(\app{E}{e},vs) &= \chi'(E,vs)\\
\chi'(\app{v}{E},vs) &= \chi'(E,vs)\\
\chi'(\wcm{E}{e},vs) &= \chi'(E,vs)\\
\chi'(\hole,vs)      &= vs
\end{align}
This definition formalizes the intuition given earlier regarding the behavior of \scheme'ccm'. Previously, we saw that the value of a \scheme'ccm' directive was a list; however, lists do not strictly exist as values in the \lc, nor in our extension. Instead of adding lists as primitive values to \cm, we employ Church-encodings to represent them natively \cite{barendregt1984lambda}.

%The accumulator-passing style definition of $\chi$ has the effect of ordering the marks from most- to least-recently placed.

With the source and target languages formally specified, we can now examine a language transformation in earnest.

@@

As transformations, desugarings put the meaning of one construct in terms of another. If the meaning of the former was given only informally, the desugaring offers a new definition as formal as the definition of the latter. In this way, desugarings offer a convenient way to provide meaning to a construct and avoid enlarging the language. If, however, the sugared language already has a formal meaning, as does \cm, the desugaring must preserve it.

We term our transformation from \cm\ to \lv\ $\mathcal{C}$, as in \emph{compile}, since we are, in essence, compiling away continuation marks. The transformation $\mathcal{C}$ is defined over each syntactic form of \cm. In order to preserve the meaning of \cm, $\mathcal{C}$ must commute with evaluation. More precisely, the following property should hold for programs $p\in\lambda_{cm}$:
\[
\begin{array}{ccc}
p & \cmrrs & v\\
\downarrow_\mathcal{C} & & \downarrow_\mathcal{C}\\
\C{p} & \lvrrs & \C{v}
\end{array}
\]
or, succinctly,
\begin{equation}
\label{meaning-preservation-property}
p\cmrrs v\implies\C{p}\lvrrs\C{v}
\end{equation}

The essence of \cm\ is that programs can apply information to and observe information about the context in which they are evaluated. Programs in \lv\ have no such facility. We can simulate this facility by explicitly passing contextual information to each term as it is evaluated. We can define $\mathcal{C}$ to transform \scheme'wcm' directives to manipulate this information and \scheme'ccm' directives to access it. In other words, we can transform \cm\ programs to mark-passing style.

However, marks alone do not account for the tail-call behavior specified by rule \ref{wcm-collapse}. Since tail-call behavior is observable (if indirectly) to \cm\ programs, we must also provide to each term information about the position in which it is evaluated. Specifically, each \scheme'wcm' directive must be

can frame this information in terms of the evaluation context and

This realization guides the intuition behind the transform: to propagate contextual information to each expression, manipulating it appropriately as the context changes.
